Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-09-18 docs/manuscript/paper.tex
some paper corrections, ldunn

\subsubsection{Measurement equation}
In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave. They are described by the geometric object $\boldsymbol{p}$ which we identify  as the 4-momentum of the radio photon. The presence of a GW induces a shift in the temporal component of the covariant 4-momentum, while the photon travels from the emitter to the observer, i.e. $\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $. One obtains \citep[e.g.][]{Maggiore}
\begin{equation}
 \Delta p_t =  \frac{\pi f_{\rm p} h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{1 + \boldsymbol{n}\cdot \boldsymbol{q} }  \left[1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right] \ ,
	\label{eq:momentum_shift}
\end{equation}
where $f_{\rm p}$ is the pulse frequency measured in the momentarily comoving reference frame of an observer. We write $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar. We take the pulsar location to be constant i.e.  neither $\boldsymbol{q}$ nor $d$ are functions of time. In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB. The barycentering correction is typically applied when generating TOAs, e.g. with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series. Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$. After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames,
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm observer} = (1,0,0,0) \ ,
\end{equation}
Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d^{(n)}) g^{(n)}(t) +  \varepsilon^{(n)}\ .
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies: 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'}(t - t') \ .	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{equation}
	g^{(n)}(t) = 1 - \frac{h_{ij} (t; \boldsymbol{x}= 0)[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \left \{1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right \} \ ,
	\label{eq:g_func}
\end{equation}
and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$. It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.

@@

\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
\begin{figure}
	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
	\caption{Spatial distribution of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work. The pulsar distances relative to the observer are also indicated, with the majority of pulsars having a distance $\sim 1$ kpc. The grey dashed line denotes the Galactic plane.}
	\label{fig:pulsar_distrib}
\end{figure}
We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}. The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA. For each pulsar we adopt the fiducial NANOGrav values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter. A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility. The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}. The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist. The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency} and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}. In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$. The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
\begin{align}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
\end{align}
where $T_{\rm cad}$ is the cadence of the timing observations and MSPs have $\ln \alpha_1 \approx -20 $, $\alpha_2 \approx 1$, $\alpha_3 \approx 2$ and $\alpha_4 \approx 2.4$. Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$. In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
\begin{eqnarray}
	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-1} \ . \label{eq:sigmap_f}
\end{eqnarray}
For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 4.28 \times 10^{-21} $ s$^{-5/2}$, $\min [ \sigma^{(n)} ] = 1.3 \times 10^{-23}$s$^{-5/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.0 \times 10^{-16}$ s$^{-5/2}$ for PSR J1939+2134 \newline

As a cross-check, we calculate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,

@@

Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct; Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise. For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$. The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$. Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred. Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar. \newline

@@

The pulsar distances $d^{(n)}$ are typically not as well constrained as $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with uncertainties typically on the order of $\sim 10 \%$ \citep{Arzoumanian2018ApJS..235...37A,Yao2017}. However for this work, whilst the pulsar distance is used when generating synthetic data, as is discussed in Section \ref{sec:parameter_estim} the pulsar distance is not used for our inference model and so we do not need to set a prior on $d^{(n)}$. Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in. That is, for $T_{\rm obs}$ =10 years the solution of Equation \eqref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied. It is therefore sufficient to consider $\gamma^{(n)}$ to be known \textit{a priori} and set it at its true injected value. We briefly explored setting an uninformative prior on $\gamma^{(n)}$ over e.g. LogUniform($10^{-15}, 10^{-10}$) s$^{-1}$ as well as setting $\gamma^{(n)}$ at some fixed value away from the true injected value (e.g. set $\gamma^{(n)} = 10^{-14} \text{s}^{-1}$ rather than $10^{-13} \text{s}^{-1}$) but the results are unchanged. Regarding $\sigma^{(n)}$, the majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22} \text{ s}^{-5/2}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}. For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$ $\text{s}^{-5/2}$). The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm em}^{(n)} (t_1)$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16} \text{s}^{-5/2}$. For our artificial dataset we set $\sigma = 10^{-21} \text{s}^{-5/2}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a single constrained prior. \newline

Since we are not setting priors on $\gamma^{(n)}$ or $d^{(n)}$ this reduces the dimensionality of the parameter space to $7 + 3N$. We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space, c.f. Equation \ref{eq:psrparams}. Explicitly, 
\begin{equation}
	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ .
\end{equation}
All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
\begin{table*}
	\centering
%	\resizebox{\textwidth}{!}{%
%	\renewcommand{\arraystretch}{1.0} % Default value: 1
	\begin{tabular}{lccll}
		\toprule
		&Parameter & Injected Values & Units & Prior  \\
		\hline
		\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
	  & $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
	  & $\delta$              & $1.0$  & radians & Uniform($-\pi/2, \pi/2$) \\
	  & $\psi$              & $2.50$ & radians & Uniform($0, 2 \pi $) \\
	  & $\Phi_0$          & $0.20$ & radians & Uniform($0, 2 \pi $) \\
	  & $h_0$            & $5 \times 10^{-15}$ & - & LogUniform($10^{-15}$, $10^{-9}$) \\
	  & $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
		\hline
		\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left( f_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{f} \right)$ \\
		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left( \dot{f}_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{\dot{f}} \right)$ \\
		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-5/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & - \\
		\bottomrule
	\end{tabular}
	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the choice of prior used for Bayesian inference on each parameter. The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}. The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}. The quantities $\epsilon^{(n)}_{f}$ and $\epsilon^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.}
	\label{tab:parameters_and_priors}
\end{table*}

@@

\section{Kalman recursion equations} \label{sec:kalman}
The linear Kalman filter operate on temporally discrete, noisy measurements $\boldsymbol{Y}_k$ which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$ via a linear transformation:
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step. The Kalman filter takes the underlying states to evolve according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the input or control matrix. and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise with covariance $\boldsymbol{Q}_k$. \newline

>>>>

\subsubsection{Measurement equation}
In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave. They are described by the geometric object $\boldsymbol{p}$ which we identify  as the 4-momentum of the radio photon. The presence of a GW induces a shift in the temporal component of the covariant 4-momentum, while the photon travels from the emitter to the observer, i.e. $\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $. One obtains \citep[e.g.][]{Maggiore}
\begin{equation}
 \Delta p_t =  \frac{\pi f_{\rm p} h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{1 + \boldsymbol{n}\cdot \boldsymbol{q} }  \left[1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right] \ ,
	\label{eq:momentum_shift}
\end{equation}
where $f_{\rm p}$ is the pulse frequency measured in the momentarily comoving reference frame of an observer. We write $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar. We take the pulsar location to be constant i.e.  neither $\boldsymbol{q}$ nor $d$ are functions of time. In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB. The barycentering correction is typically applied when generating TOAs, e.g. with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series. Some pulsars, including some PTA pulsars, do have non-negligible proper motions $\mathcal{O} (100)$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper. Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$. After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames,
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm observer} = (1,0,0,0) \ ,
\end{equation}
Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d^{(n)}) g^{(n)}(t) +  \varepsilon^{(n)}\ .
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies: 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'}(t - t') \ .	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{equation}
	g^{(n)}(t) = 1 - \frac{h_{ij} (t; \boldsymbol{x}= 0)[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \left \{1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right \} \ ,
	\label{eq:g_func}
\end{equation}
and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$. It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.

@@

\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
\begin{figure}
	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work. The pulsar distances relative to the observer are also indicated, with the majority of pulsars having a distance $\sim 1$ kpc. The grey dashed line denotes the Galactic plane.}
	\label{fig:pulsar_distrib}
\end{figure}
We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}. The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA. For each pulsar we adopt the fiducial NANOGrav values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter. A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility. The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}. The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist. The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency} and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}. In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$. The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
\begin{align}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
\end{align}
where $T_{\rm cad}$ is the cadence of the timing observations and MSPs have $\ln \alpha_1 \approx -20 $, $\alpha_2 \approx 1$, $\alpha_3 \approx 2$ and $\alpha_4 \approx 2.4$. Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$. In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
\begin{eqnarray}
	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-1} \ . \label{eq:sigmap_f}
\end{eqnarray}
For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 4.28 \times 10^{-21} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.3 \times 10^{-23}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.0 \times 10^{-16}$ s$^{-3/2}$ for PSR J1939+2134 \newline

As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,

@@

Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct; Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise. For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz. The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$ Hz. Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred. Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar. \newline

@@

The pulsar distances $d^{(n)}$ are typically not as well constrained as $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with uncertainties typically on the order of $\sim 10 \%$ \citep{Arzoumanian2018ApJS..235...37A,Yao2017}. However for this work, whilst the pulsar distance is used when generating synthetic data, as is discussed in Section \ref{sec:parameter_estim} the pulsar distance is not used for our inference model and so we do not need to set a prior on $d^{(n)}$. Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in. That is, for $T_{\rm obs}$ =10 years the solution of Equation \eqref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied. It is therefore sufficient to consider $\gamma^{(n)}$ to be known \textit{a priori} and set it at its true injected value. We briefly explored setting an uninformative prior on $\gamma^{(n)}$ over e.g. LogUniform($10^{-15}, 10^{-10}$) s$^{-1}$ as well as setting $\gamma^{(n)}$ at some fixed value away from the true injected value (e.g. set $\gamma^{(n)} = 10^{-14} \text{s}^{-1}$ rather than $10^{-13} \text{s}^{-1}$) but the results are unchanged. Regarding $\sigma^{(n)}$, the majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22} \text{ s}^{-3/2}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}. For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$ $\text{s}^{-3/2}$). The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm em}^{(n)} (t_1)$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16} \text{s}^{-3/2}$. For our artificial dataset we set $\sigma = 10^{-21} \text{s}^{-3/2}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a single constrained prior. \newline

Since we are not setting priors on $\gamma^{(n)}$ or $d^{(n)}$ this reduces the dimensionality of the parameter space to $7 + 3N$. We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space, c.f. Equation \ref{eq:psrparams}. Explicitly, 
\begin{equation}
	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ .
\end{equation}
All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
\begin{table*}
	\centering
%	\resizebox{\textwidth}{!}{%
%	\renewcommand{\arraystretch}{1.0} % Default value: 1
	\begin{tabular}{lccll}
		\toprule
		&Parameter & Injected Values & Units & Prior  \\
		\hline
		\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
	  & $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
	  & $\delta$              & $1.0$  & radians & Uniform($-\pi/2, \pi/2$) \\
	  & $\psi$              & $2.50$ & radians & Uniform($0, 2 \pi $) \\
	  & $\Phi_0$          & $0.20$ & radians & Uniform($0, 2 \pi $) \\
	  & $h_0$            & $5 \times 10^{-15}$ & - & LogUniform($10^{-15}$, $10^{-9}$) \\
	  & $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
		\hline
		\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left( f_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{f} \right)$ \\
		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left( \dot{f}_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{\dot{f}} \right)$ \\
		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & - \\
		\bottomrule
	\end{tabular}
	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the choice of prior used for Bayesian inference on each parameter. The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}. The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}. The quantities $\epsilon^{(n)}_{f}$ and $\epsilon^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.}
	\label{tab:parameters_and_priors}
\end{table*}

@@

\section{Kalman recursion equations} \label{sec:kalman}
The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation:
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step. The Kalman filter takes the underlying states to evolve according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the control matrix. $\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise with covariance $\boldsymbol{Q}_k$. \newline
